---
title: "METABRIC Discovery Basal-like Analysis"
author: "Christopher Conley, Pei Wang, Umut Ozbek, Jie Peng"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(message=F, warning=F, cache = TRUE)
```

This document highlights the full network analysis pipeline of spaceMap applied to 86 basal-like breast cancer samples. There are three major sections, namely (i) model tuning with cross validation, (ii) comparisons of CV.Vote and Boot.Vote ensemble networks, (iii) and network analysis efforts. Model tuning is an iterative process of arriving at the final tuning parameters, which we detail with various diagnostic tools. 

## CV Vote Model Tuning 

The following function helps consolidate the diagnostic information for different iterations of CV.

```{r}
source("dxCV.R")
```

### Tuning iteration 1

This first tuning iteration of spaceMap shows CV-score profiles that suggest to shift to lower the tuning parameter values for `lam1,lam2,lam3`. 

```{r}
dx1 <- dxsmap(file = "~/scratch-data/neta-metabric/disc-cv-vote/basal/01/basal-01.rda", F)
```


### Tuning iteration 2

The second tuning iteration does not bring much clarity to optimal neighborhood for `lam1`. However, it would appear that near optimal neighborhoods for `lam2,lam3` have been found. 

```{r}
dx2 <- dxsmap(file = "~/scratch-data/neta-metabric/disc-cv-vote/basal//02/basal-02.rda", T, c(0,0.5))
```

### Tuning iteration 3

Finally in the third iteration we see that there are several `lam1` that produce almost equally good CV-scores and the valley is relatively shallow. In this iteration, we feel fairly confident still that the optimal neighborhood is centered at roughly `lam1=67, lam2=27, lam3=10`. 

```{r}
dx3 <- dxsmap(file = "~/scratch-data/neta-metabric/disc-cv-vote/basal/03/basal-03-dropNA.rda", T, c(0,0.2))
```

### Tuning iteration 4

The fourth iteration produces CV-score profiles suggesting we have found near optimal values for all three tuning parameters.

```{r}
dx4 <- dxsmap(file = "~/scratch-data/neta-metabric/disc-cv-vote/basal/04/basal-04-dropNA.rda", T, c(0,0.2))
```

The following plot demonstrates the consistent stability of the CV-selected tuning parameters across different tuning iterations; labels 2,3,4 denote the CV-selected tuning parameters for each iteration; the background points correspond to tuning grids for each iteration (2=dots,3=squares,4=pluses). None of the tuning parameters are on the boundary. 

```{r, echo=FALSE}
i02 <- new.env()
load("~/scratch-data/neta-metabric/disc-cv-vote/basal/02/basal-02.rda", envir = i02)
tmap2 <- i02$tmap
i03 <- new.env()
load("~/scratch-data/neta-metabric/disc-cv-vote/basal/03/basal-03-dropNA.rda", envir = i03)
tmap3 <- i03$tmap
i04 <- new.env()
load("~/scratch-data/neta-metabric/disc-cv-vote/basal/04/basal-04-dropNA.rda", envir = i04)
tmap4 <- i04$tmap

par(mfrow = c(1,2))
plot(tmap2$lam1, tmap2$lam2, ylab = "lam2", xlab = "lam1", 
     pch = 19, xlim = c(60,79), ylim = c(14,35))
points(tmap3$lam1, tmap3$lam2, pch = 22)
points(tmap4$lam1, tmap4$lam2, pch = "+")
points(i02$cvsmap$minTune$lam1, i02$cvsmap$minTune$lam2, pch = "2", cex = 2)
points(i03$cvsmap$minTune$lam1, i03$cvsmap$minTune$lam2, pch = "3", cex = 2)
points(i04$cvsmap$minTune$lam1, i04$cvsmap$minTune$lam2, pch = "4", cex = 2)


plot(tmap2$lam2, tmap2$lam3, ylab = "lam3", xlab = "lam2", 
     pch = 19, xlim = c(14,35), ylim = c(0, 80))
points(tmap3$lam2, tmap3$lam3, pch = 22)
points(tmap4$lam2, tmap4$lam3, pch = "+")
points(i02$cvsmap$minTune$lam2, i02$cvsmap$minTune$lam3, pch = "2", cex = 2)
points(i03$cvsmap$minTune$lam2, i03$cvsmap$minTune$lam3, pch = "3", cex = 2)
points(i04$cvsmap$minTune$lam2, i04$cvsmap$minTune$lam3, pch = "4", cex = 2)
```


The best CV-score of iteration 2 (score: `r round(i04$cvsmap$logcvScore,3)`) is lower than iteration 3 (score: `r round(i03$cvsmap$logcvScore,3)`) and iteration 3 (score: `r round(i02$cvsmap$logcvScore,3)`). We move onto Boot.Vote analysis with CV-selected parameters from the 2nd iteration of tuning. 

## Boot.Vote analysis

The drop in edge count from the training folds to the CV.Vote model and the Boot.Vote model are indications of model instability. The large drop in model size for the the Boot.Vote model (B = 250) is likely due to the limited sample size and signal issues of the data.

```{r}
bv <- readRDS(file.path("~/scratch-data/neta-metabric/disc-boot-vote/basal/01/","basal-boot-vote-01.rds"))
dbv <- data.frame(xy = nonZeroWhole(bv$bv$xy,0.0),
                  yy = nonZeroUpper(bv$bv$yy,0.0),
                  type = "Boot.Vote")
dx2$gcvdrop + geom_point(data = dbv, size  = 2)
```

## Gene Ontology annotation

The identifier is already an Entrez accession for the mRNA features. Obtain the Entrez accessions with their GO mappings.

```{r}
library(Biobase)
filterdir <- "~/scratch-data/neta-metabric/disc-filtered/"
rnaset <- readRDS(file = file.path(filterdir,"disc-mrna-eset-nostd-union-dropout-std-Basal.rds"))
rna_entrez_ids <- pData(featureData(rnaset))[,"EntrezReannotated"]
library(IRanges)
library(org.Hs.eg.db)
hs_feature_list <- function(db, eg) { 
  mk <- mappedkeys(db)
  xx <- as.list(db[eg[eg %in% mk]])
}
leg2go <- hs_feature_list(db = org.Hs.egGO, eg = rna_entrez_ids)
```

There are `r sum(is.na(rna_entrez_ids))` Entrez accesions without any GO mapping. Map the Entrez accessions to GO biological process terms with some valid evidence (not “ND”).

```{r}
library(foreach)
leg2go_filtered <- foreach(got = leg2go, eg = names(leg2go)) %do% { 
  bp_domain <- sapply(got, function(x) x[["Ontology"]]) == "BP"
  not_nd_domain <- sapply(got, function(x) x[["Evidence"]]) != "ND"
  bp_got <- got[bp_domain & not_nd_domain]
  sapply(bp_got, function(x) x[["GOID"]]) 
}
names(leg2go_filtered) <- names(leg2go)
leg2go_filtered <- leg2go_filtered[sapply(leg2go_filtered, length) != 0]
```

Reverse the mapping so that the key is a GO biological process and the values are a character vector of Entrez gene accessions. Keep those keys that have at least 15 but no more than 300 RefSeq accessions.

```{r}
suppressPackageStartupMessages(library(topGO))
go2eg <- inverseList(leg2go_filtered)
#remove duplicate Entrez genes
go2eg <- lapply(go2eg, function(x) x[!duplicated(x)])
gosize <- sapply(go2eg, length)
go2eg <- go2eg[ gosize >= 15 & gosize <= 300]
```

There are `r length(go2eg)` GO terms satisfying these conditions. Assure each go term has a list of non-duplicate ids and save the GO background.

```{r}
stopifnot(all(sapply(go2eg, anyDuplicated) == 0))
saveRDS(go2eg, file = file.path(filterdir, "basal-go-bp-to-entrez-gene-list.rds"))
```

Obtain human readable versions of the GO accessions for reporting GO enrichment later. 

```{r}
library(AnnotationDbi)
#human readable 
process_alias <- Term(names(go2eg))
```

## Network Analysis on Boot.Vote

Read in the probe and CNA feature information stored in the basal-like ExpressionSet.

```{r}
library(Biobase)
yinfo <- pData(featureData(readRDS(file = file.path(filterdir,"disc-mrna-eset-nostd-union-dropout-std-Basal.rds"))))
xinfo <- pData(featureData(readRDS(file = file.path(filterdir,"disc-cna-eset-nostd-union-dropout-multi-std-Basal.rds"))))
yinfo$start <- as.integer(yinfo$start)
yinfo$end <- as.integer(yinfo$end)
xinfo$start <- as.integer(xinfo$start)
xinfo$end <- as.integer(xinfo$end)
```

### Collapse probes to gene level

For the network analysis, we decided to collapse probe information down onto the gene-level. For example, if two probes map to the same locus, we take the union of their edges to form a single node. First, import and label fetures of the CV.Vote network. 

```{r}
net <- bv$bv
rownames(net$xy) <- xinfo$id; colnames(net$xy) <- yinfo$id;
rownames(net$yy) <- yinfo$id; colnames(net$yy) <- yinfo$id;
bdeg <- bv$bdeg
colnames(bdeg$xy) <- xinfo$id; colnames(bdeg$yy) <- yinfo$id;
```

The following code merges probes onto the gene level. 

```{r}
egtab <- table(yinfo$EntrezReannotated)
eg2probeid <- lapply(names(egtab), function(eg) {
  yinfo$id[yinfo$EntrezReannotated %in% eg]  
})
#take union of edges between probes with same Entrez gene id
meg <- eg2probeid[sapply(eg2probeid, length) > 1]
for(m in meg) { 
  uyedge <- apply(X = net$yy[m,], MARGIN = 2, FUN = max)
  uxedge <- apply(X = net$xy[,m], MARGIN = 1, FUN = max)
  net$yy[m[1],] <- uyedge
  net$xy[,m[1]] <- uxedge
}
#drop secondary listing of probes
secprobes <- unlist(sapply(meg, function(m) m[2:length(m)]))
keepers <- setdiff(rownames(net$yy), secprobes)
net$yy <- net$yy[keepers,keepers]
net$xy <- net$xy[,keepers]
#adjust order of yinfo
yinfo <- yinfo[match(keepers,yinfo$id),]
bdeg$yy <- bdeg$yy[,keepers]
```

Next we replace the probe ID with the Entrez gene id. 

```{r}
probe2eg <- yinfo[,c("id", "EntrezReannotated")]
#all duplicates should be missing values now. 
eg <- probe2eg$EntrezReannotated
stopifnot(all(is.na(eg[which(duplicated(eg))])))
#replace ID with Entrez gene
yinfo$probeid <- yinfo$id
yinfo$id <- ifelse(is.na(eg), yinfo$id, eg)
stopifnot(!anyDuplicated(yinfo$id))
```

Update the feature name labels of the network after merging the probes to the gene level. 

```{r}
rownames(net$xy) <- xinfo$id; colnames(net$xy) <- yinfo$id;
rownames(net$yy) <- yinfo$id; colnames(net$yy) <- yinfo$id;
colnames(bdeg$xy) <- xinfo$id; colnames(bdeg$yy) <- yinfo$id; 
```

Now there are `r length(secprobes)` mRNA feature nodes lost from merging probes, leaving `r length(keepers)` mRNA gene nodes. Import the network to igraph. 

```{r}
library(spacemap)
ig <- spacemap::adj2igraph(yy = net$yy, xy = net$xy, yinfo = yinfo, xinfo = xinfo)
```

### Hub analysis

Rank the CNA- and mRNA- hubs hubs according to their (in + out) degree. <!--the average degree rank. Accordingly, the most highly ranked hubs will have the most consistently high degree across the network ensemble.-->

```{r}
ig <- rankHub(ig = ig, level = "y")
ig <- rankHub(ig = ig, level = "x")
```

Next label CNA--mRNA edges as being regulated in cis or in trans. 

```{r}
library(IRanges)
ig <- cisTrans(ig = ig, level = "x-y", cw = 2e6)
```

We then report CNA-hubs with their cis genes. 

```{r}
xhubs <- reportHubs(ig, top = 100, level = "x")
```

```{r, echo=FALSE}
knitr::kable(xhubs, row.names = FALSE)
```

We will want these results in LaTex format as well.  Import the utility functions to write these results for the LaTex format. 

```{r}
source("tab2tex.R")
netadir <- "~/scratch-data/neta-metabric/disc-neta/"
xhubs2textab(xhubs = xhubs, 
             cap = "CNA-hubs of basal-like subtype discovery samples and ranked by out degree.", 
             tfile = file.path(netadir, "basal-bv-p90-01-xhubs-deg.tex"))
#under 
tmp <- rankHub(ig = ig, bdeg = bdeg$xy, level = "x")
xhubs2textab(xhubs =  reportHubs(tmp, top = 100, level = "x"), 
             cap = "CNA-hubs of basal-like subtype discovery samples and ranked by mean rank of degree.", 
             tfile = file.path(netadir, "basal-bv-p90-01-xhubs-bdeg.tex"))
```

Report what are the stable mRNA-hub genes: 

```{r}
yhubs <- reportHubs(ig, top = 50, level = "y")
```

```{r, echo=FALSE}
knitr::kable(yhubs, row.names = FALSE)
```

Obtain the GO-neighbor percentages for CNA hubs. 

```{r}
hgp <- xHubEnrich(ig = ig, go2eg = go2eg)
hgp2 <- hgp[hgp$neighbor_percentage > 50 & hgp$degree > 7,]
```

Report the larger CNA-hubs (degree > 7) with with at least 50% GO-neighbor percentage. 

```{r, echo=FALSE}
knitr::kable(hgp2)
```

### Module analysis

Use cluster edge betweenness to find modules. 

```{r}
library(igraph)
mods <- cluster_edge_betweenness(ig)
```

Test for module Enrichment with the hypergeometric test. 

```{r}
outmod <- modEnrich(ig = ig, mods = mods, levels = "y", go2eg = go2eg, process_alias = process_alias, prefix = "R")
```

Report the significantly enriched GO terms organized by module.

```{r, eval = FALSE}
outmod$etab
```

```{r, echo=FALSE}
knitr::kable(outmod$etab, row.names = FALSE)
```

We export the module-GO enrichment results in LaTex format. 

```{r}
mod2textab(etab = outmod$etab, 
           cap ="Enrichment of GO terms in network modules of basal-like subtype discovery samples (ranked by most significant)", 
           tfile = file.path(netadir, "basal-bv-p90-01-etab.tex"))
```

## Save results

```{r}
filename <- file.path(netadir, "basal-boot-vote-01.graphml")
#delete nodes without edges from the graph object
vis <- delete_vertices(graph = outmod$ig, v = V(outmod$ig)[igraph::degree(outmod$ig) == 0])
igraph::write_graph(graph = vis, file = filename, format = "graphml")
save.image(file = file.path(netadir, "basal-full-analysis.rda"))
```
